msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-12 paper.tex
Updates for review.

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

@@

For our purposes here, an ARG is a realisation of the coalescent with
recombination, in the Griffiths sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

@@

One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
(encoded using the representation of the \texttt{record\_full\_arg} option)
and mutational patterns under
the standard coalescent and infinite sites mutation model.

% We provide two functions to facilitate likelihood-based inference.
% Both are implemented only for the simplest case of the standard ARG with a
% constant population size, and require tree sequences compatible with the
% \texttt{record\_full\_arg} option as their arguments.
% \begin{itemize}
% \item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
% the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
% per-generation recombination probability \texttt{r} and population size \texttt{N}.
% Specifically, the function returns the logarithm of
% \begin{equation*}
% \Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
% 	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
% 		+ r l_i \Big] t_i  \Bigg),
% \end{equation*}
% where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
% $k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
% in that interval that would split ancestral material should they recombine,
% $q$ is the total number of events in the tree sequence $\texttt{ts}$,
% $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
% of time intervals which end in a recombination, and $g_i$ is the corresponding
% \emph{gap}:  the length of material in which the recombination at the end of the
% $i$th period could have taken place without affecting the resulting genealogy.
% A recombination in ancestral material will always have $g_i = 1$, while a recombination
% in trapped non-ancestral material can result in other values of $g_i$.
% For a continuous model of the genome and a recombination in ancestral material,
% we set $g_i = 1$ and interpret the result as a density.
% \item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
% natural logarithm of the probability of the mutations recorded in the tree sequence
% \texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
% a normalising constant which depends on the pattern of mutations,
% but not on the tree sequence or the per-site, per-generation mutation
% probability \texttt{m}.
% Specifically, the function returns the logarithm of
% \begin{equation*}
% e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
% \prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
% \end{equation*}
% where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
% in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
% total branch length on which $\gamma$ could have arisen while appearing on all
% of the leaves of \texttt{ts} it does, and on no others.
% Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
% mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
% edges.
% \end{itemize}

@@

\appendix
\label{app-multiple-mergers}
\section{Multiple merger coalescent model}

>>>>

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows.
We may consider the big ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral material and
consequently wait an exponentially large time until all sampled lineages
coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
which we only track genetic ancestors, and wait until all marginal trees
have fully coalesced (like Hudson's algorithm).

@@

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

@@

One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
(encoded using the representation of the \texttt{record\_full\_arg} option)
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details.

@@

\appendix

\label{app-likelihoods}
\section{Likelihood calculations}

We provide two functions to facilitate likelihood-based inference.
Both are implemented only for the simplest case of the standard ARG with a
constant population size, and require tree sequences compatible with the
\texttt{record\_full\_arg} option as their arguments.
\begin{itemize}
\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
per-generation recombination probability \texttt{r} and population size \texttt{N}.
Specifically, the function returns the logarithm of
\begin{equation*}
\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
		+ r l_i \Big] t_i  \Bigg),
\end{equation*}
where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
in that interval that would split ancestral material should they recombine,
$q$ is the total number of events in the tree sequence $\texttt{ts}$,
$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
of time intervals which end in a recombination, and $g_i$ is the corresponding
\emph{gap}:  the length of material in which the recombination at the end of the
$i$th period could have taken place without affecting the resulting genealogy.
A recombination in ancestral material will always have $g_i = 1$, while a recombination
in trapped non-ancestral material can result in other values of $g_i$.
For a continuous model of the genome and a recombination in ancestral material,
we set $g_i = 1$ and interpret the result as a density.
\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
natural logarithm of the probability of the mutations recorded in the tree sequence
\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
a normalising constant which depends on the pattern of mutations,
but not on the tree sequence or the per-site, per-generation mutation
probability \texttt{m}.
Specifically, the function returns the logarithm of
\begin{equation*}
e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
\end{equation*}
where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
total branch length on which $\gamma$ could have arisen while appearing on all
of the leaves of \texttt{ts} it does, and on no others.
Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
edges.
\end{itemize}

\label{app-multiple-mergers}
\section{Multiple merger coalescent model}
